---
title: "TIME SERIES FINAL PROJECT"
author: "Jani Shariff Shaik"
date: "2024-12-13"
output: word_document
---
```{r}
# Load required libraries
library(ggplot2)
library(forecast)
library(tseries)
library(dplyr)
library(zoo)
library(lubridate)


# Load the dataset
df <- read.csv("C:/Users/janis/OneDrive/Desktop/Final Project/Time Series/renewable_energy_monthly_data.csv")

# Convert 'Date' to Date format
df$Date <- as.Date(df$Date, format = "%d-%m-%Y")

# Set Date as the index
df <- df %>% arrange(Date)  # Ensure data is sorted by Date

# Preview the dataset
head(df)
data<-df

```
```{r}
str(data)
summary(data)

```
```{r}
total_nulls <- sum(is.na(data))
print(total_nulls)
```
HANDLING OUTLIERS
```{r}
# Calculate the Interquartile Range (IQR)
Q1 <- quantile(data$Energy_Production_MWh, 0.25)
Q3 <- quantile(data$Energy_Production_MWh, 0.75)
IQR <- Q3 - Q1

# Define the outlier bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
outliers <- data$Energy_Production_MWh[data$Energy_Production_MWh < lower_bound | 
                                       data$Energy_Production_MWh > upper_bound]

# Count the number of outliers
num_outliers <- length(outliers)
cat("Number of Outliers:", num_outliers, "\n")

```
```{r}
# Boxplot to visualize outliers
ggplot(data, aes(y = Energy_Production_MWh)) +
  geom_boxplot(fill = "blue", color = "black", outlier.color = "red", outlier.size = 2) +
  labs(title = "Boxplot of Energy Production ", 
       y = "Energy Production (MWh)") +
  theme_minimal()

```
3. Exploratory Data Analysis
3.1 Summary Statistics and Data Visualization
```{r}
# Summary statistics
summary(data$Energy_Production_MWh)

# Plot energy production over time
ggplot(data, aes(x = Date, y = Energy_Production_MWh)) +
  geom_line(color = "blue") +
  labs(title = "Energy Production Over Time", x = "Date", y = "Energy Production (MWh)") +
  theme_minimal()

# Histogram of energy production
ggplot(data, aes(x = Energy_Production_MWh)) +
  geom_histogram(binwidth = 500, fill = "blue", color = "black") +
  labs(title = "Distribution of Energy Production", x = "Energy Production (MWh)", y = "Frequency") +
  theme_minimal()

```

3.2 Seasonality Strength Calculation
```{r}
data$Date <- as.Date(data$Date, format = "%d-%m-%Y")
data <- data %>% arrange(Date)
energy_ts <- ts(data$Energy_Production_MWh, 
                start = c(year(min(data$Date)), month(min(data$Date))), 
                frequency = 12)
# Decompose the time series (multiplicative)
decomposed <- decompose(energy_ts, type = "multiplicative")

# Seasonality strength
seasonality_strength <- var(decomposed$seasonal, na.rm = TRUE) /
                        var(data$Energy_Production_MWh, na.rm = TRUE)
cat("Seasonality Strength:", seasonality_strength, "\n")

```
```{r}
head(energy_ts)
```

3.3 Visualize Rolling Statistics
```{r}
# Calculate rolling mean and rolling standard deviation
rolling_mean <- rollmean(data$Energy_Production_MWh, k = 12, fill = NA) # 12-month rolling mean
rolling_std <- rollapply(data$Energy_Production_MWh, width = 12, FUN = sd, fill = NA) # 12-month rolling std

# Plot rolling statistics
ggplot() +
  geom_line(aes(x = data$Date, y = data$Energy_Production_MWh), color = "blue") +
  geom_line(aes(x = data$Date, y = rolling_mean), color = "red") +
  geom_line(aes(x = data$Date, y = rolling_std), color = "black") +
  labs(title = "Rolling Mean & Standard Deviation", x = "Date", y = "Energy Production (MWh)") +
  theme_minimal()


```
Stationarity Testing and Transformation
```{r}
# Augmented Dickey-Fuller test
adf_test <- adf.test(energy_ts, alternative = "stationary")
cat("ADF Test p-value:", adf_test$p.value, "\n")

# Log transformation if variance increases with trend
log_energy_ts <- log(energy_ts)

# Differencing if non-stationary
diff_energy_ts <- diff(log_energy_ts)
plot(diff_energy_ts, main = "Differenced Time Series")

```
plots
```{r}
# Plot original time series
plot(energy_ts, main = "Original Time Series", ylab = "Energy Production (MWh)", xlab = "Time", col = "blue")

# Plot log-transformed time series
plot(log_energy_ts, main = "Log-Transformed Time Series", ylab = "Log of Energy Production (MWh)", xlab = "Time", col = "blue")

# Plot differenced time series
plot(diff_energy_ts, main = "Differenced Time Series", ylab = "Differenced Log of Energy Production", xlab = "Time", col = "red")

```
ACF AND PACF
```{r}
# Original time series
plot(energy_ts, main = "Original Time Series", ylab = "Energy Production (MWh)", xlab = "Time", col = "blue")

# Log-transformed time series
log_energy_ts <- log(energy_ts)
plot(log_energy_ts, main = "Log-Transformed Time Series", ylab = "Log of Energy Production (MWh)", xlab = "Time", col = "green")

# Differenced time series
diff_energy_ts <- diff(log_energy_ts)
plot(diff_energy_ts, main = "Differenced Time Series", ylab = "Differenced Log of Energy Production", xlab = "Time", col = "red")

# ACF and PACF for differenced data
acf(diff_energy_ts, main = "ACF of Differenced Series")
pacf(diff_energy_ts, main = "PACF of Differenced Series")

```
ACF AND PACF
```{r}
# Plot ACF with proper lags
acf(diff_energy_ts, lag.max = 24, main = "ACF of Differenced Series", xaxt = "n") # Lag.max specifies maximum lag
axis(1, at = seq(0, 24, by = 1)) # Customize x-axis labels to display actual lags

# Plot PACF with proper lags
pacf(diff_energy_ts, lag.max = 24, main = "PACF of Differenced Series", xaxt = "n") # Lag.max specifies maximum lag
axis(1, at = seq(0, 24, by = 1)) # Customize x-axis labels to display actual lags

```

5. Build ARIMA and SARIMA Models
5.1 Grid Search for ARIMA Parameters
```{r}
best_fit <- list()
min_aic <- Inf
for (p in 0:3) {
  for (d in 0:2) {
    for (q in 0:3) {
      model <- try(Arima(energy_ts, order = c(p, d, q)), silent = TRUE)
      if (!inherits(model, "try-error") && model$aic < min_aic) {
        min_aic <- model$aic
        best_fit <- list(p = p, d = d, q = q)
      }
    }
  }
}

# Print the best ARIMA parameters
print(paste("Best ARIMA Parameters: p =", best_fit$p, ", d =", best_fit$d, ", q =", best_fit$q))

```
5.2 Fit and Evaluate ARIMA
```{r}
# Fit ARIMA model
arima_model <- Arima(energy_ts, order = c(best_fit$p, best_fit$d, best_fit$q))
summary(arima_model)

# Residual diagnostics
checkresiduals(arima_model)

```
5.3 Fit and Evaluate SARIMA
```{r}
# Fit SARIMA model
sarima_model <- Arima(energy_ts, order = c(2, 1, 2), seasonal = c(0, 1, 1))
summary(sarima_model)

# Residual diagnostics
checkresiduals(sarima_model)

```
6. Forecasting and Evaluation
6.1 Forecast Using SARIMA
```{r}
# Split data into training and testing sets
train <- window(energy_ts, end = c(2018, 12))
test <- window(energy_ts, start = c(2019, 1))

# Forecast for the testing period
sarima_forecast <- forecast(sarima_model, h = length(test))

# Plot forecast vs actual
autoplot(sarima_forecast) +
  autolayer(test, series = "Actual") +
  labs(title = "SARIMA Forecast vs Actual", x = "Time", y = "Energy Production (MWh)") +
  theme_minimal()

```
6.2 Calculate Forecasting Accuracy
```{r}
# Perform Time Series Cross-Validation
time_series_cv <- function(ts_data, initial_window, horizon, step) {
  results <- data.frame(Fold = numeric(), RMSE = numeric(), MAE = numeric(), MAPE = numeric())
  
  for (i in seq(initial_window, length(ts_data) - horizon, by = step)) {
    train <- ts_data[1:i]
    test <- ts_data[(i + 1):(i + horizon)]
    
    # Fit SARIMA model
    sarima_model <- auto.arima(train, seasonal = TRUE)
    
    # Forecast for the horizon
    forecasted <- forecast(sarima_model, h = horizon)$mean
    
    # Calculate metrics
    rmse <- sqrt(mean((test - forecasted)^2))
    mae <- mean(abs(test - forecasted))
    mape <- mean(abs((test - forecasted) / test)) * 100
    
    results <- rbind(results, data.frame(Fold = i, RMSE = rmse, MAE = mae, MAPE = mape))
  }
  
  return(results)
}

# Apply Cross-Validation on Energy Production Time Series
initial_window <- 96  # First 8 years for training
horizon <- 12         # Forecasting horizon (12 months)
step <- 12            # Step size (1 year)

cv_results <- time_series_cv(energy_ts, initial_window, horizon, step)
print(cv_results)

# Calculate average metrics
cat("Average RMSE:", mean(cv_results$RMSE), "\n")
cat("Average MAE:", mean(cv_results$MAE), "\n")
cat("Average MAPE:", mean(cv_results$MAPE), "\n")


```

7. Forecast Future Energy Production
```{r}
library(lubridate)  # For working with date components

# Dynamic Forecast Updates Function
dynamic_forecast_update <- function(data, current_date, forecast_horizon) {
  # Extract data up to the current date
  updated_data <- data[data$Date <= current_date, ]
  
  # Convert to time series
  updated_ts <- ts(updated_data$Energy_Production_MWh, 
                   start = c(year(min(updated_data$Date)), month(min(updated_data$Date))), 
                   frequency = 12)
  
  # Fit SARIMA model
  sarima_model <- auto.arima(updated_ts, seasonal = TRUE)
  
  # Forecast for the horizon
  future_forecast <- forecast(sarima_model, h = forecast_horizon)
  
  # Plot the forecast
  autoplot(future_forecast) +
    labs(title = paste("Dynamic Forecast from", current_date), x = "Time", y = "Energy Production (MWh)") +
    theme_minimal()
  
  return(future_forecast)
}

# Example: Dynamic Forecast for Next 12 Months
dynamic_forecast <- dynamic_forecast_update(data, as.Date("2020-12-31"), 12)
print(dynamic_forecast)

# Plot the forecast
autoplot(dynamic_forecast) +
  labs(title = "Dynamic Forecast for Next 12 Months",
       x = "Time",
       y = "Energy Production (MWh)") +
  theme_minimal()


```
```{r}
library(lubridate)  # For working with date components
library(forecast)   # For SARIMA and forecasting
library(ggplot2)    # For enhanced plotting

# Dynamic Forecast Updates Function
dynamic_forecast_update <- function(data, current_date, forecast_horizon) {
  # Filter data up to the current date
  updated_data <- data[data$Date <= current_date, ]
  
  # Convert to time series object
  updated_ts <- ts(
    updated_data$Energy_Production_MWh,
    start = c(year(min(updated_data$Date)), month(min(updated_data$Date))),
    frequency = 12
  )
  
  # Fit SARIMA model
  sarima_model <- auto.arima(updated_ts, seasonal = TRUE)
  
  # Print the model summary
  print(summary(sarima_model))
  
  # Forecast for the specified horizon
  future_forecast <- forecast(sarima_model, h = forecast_horizon)
  
  # Convert forecast to data frame for custom plotting
  forecast_df <- data.frame(
    Date = seq(
      from = current_date + months(1),
      by = "month",
      length.out = forecast_horizon
    ),
    Point.Forecast = future_forecast$mean,
    Lo.80 = future_forecast$lower[, 1],
    Hi.80 = future_forecast$upper[, 1],
    Lo.95 = future_forecast$lower[, 2],
    Hi.95 = future_forecast$upper[, 2]
  )
  
  # Enhanced Plot
  forecast_plot <- ggplot(forecast_df, aes(x = Date)) +
    geom_line(aes(y = Point.Forecast), color = "blue", size = 1) +  # Forecasted values
    geom_ribbon(aes(ymin = Lo.80, ymax = Hi.80), fill = "lightblue", alpha = 0.3) +  # 80% confidence interval
    geom_ribbon(aes(ymin = Lo.95, ymax = Hi.95), fill = "blue", alpha = 0.2) +  # 95% confidence interval
    labs(
      title = paste("Dynamic Forecast from", current_date),
      x = "Time",
      y = "Energy Production (MWh)",
      caption = "Shaded areas represent 80% and 95% confidence intervals"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  
  # Return the forecast and plot
  return(list(forecast = future_forecast, plot = forecast_plot))
}

# Example Usage: Dynamic Forecast for Next 12 Months
# Ensure `data` contains `Date` and `Energy_Production_MWh` columns
dynamic_forecast <- dynamic_forecast_update(data, as.Date("2020-12-31"), 12)

# Print forecast details
print(dynamic_forecast$forecast)

# Display the enhanced plot
print(dynamic_forecast$plot)



```